---
title: "2018 Model Extension"
author: "Sean Davern"
date: "June 13, 2019"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---
# Load Data
Loading cleaned and transformed giving data:
```{r}
df <- readRDS("../data/Cleaned and Transformed Giving Data.rds")
df
```
# Accommodate Excluding Selected Months
12/26/2016 had an unusual single gift skewing it. What would be best would be to subtract this gift from the month's total.  I don't have it right now, so I'll just exclude the entire month.
```{r}
df <- mutate(df, excluded = FALSE)
df$excluded[as.Date(df$week.ending) == "2016-12-25"] <- TRUE
```

# Modeling
Regressing the R version of the 2013 model:
$$\textit{Monthly Giving} = a+b_{year}+c_{month}$$
```{r}
mod <- lm(
  formula = total ~
    year + month,
  data = df[df$excluded!=TRUE,]
)
```
...gives the following

# Model Predictions

```{r echo=FALSE}
library("magrittr")
library("ggplot2")
library("RColorBrewer")
cbPalette <- brewer.pal(12,"Paired")  # Used http://colorbrewer2.org/
names(cbPalette) <- levels(month.name)
colScale <- scale_color_manual(name = "month", values = cbPalette)
t.size <- 9
df$week.ending <- as.Date(df$week.ending, "%Y-%m-%d")
# The model's predicted values don't have values at the excluded months.
# Calculate those:
df <- mutate(df, predicted.values = predict(mod,df)) %>%
  mutate(residual.values = total - predicted.values )
ggplot(df, aes(x = as.Date(week.ending), y = total, color=month )) +
  geom_point() + geom_line(color='blue', 
                           data = df, 
                           aes(x=week.ending,y=predicted.values)) +
  colScale + labs(x = "Date", 
                  y = "Month's Total Giving ($)",
                  title = "Monthly Giving Model & Data") + 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") + 
  scale_y_continuous(breaks = seq(10000, 65000, by=5000),
                     labels = scales::comma) +
  theme(text = element_text(size=t.size),
        panel.grid.major = element_line(color = "azure2",
                                        linetype = "dashed",
                                        line), 
        axis.text.x = element_text(angle=45, hjust=1))
```
In this case it is far more clear that giving was increasing through 2016 which would be expected to improve the statistical significance of the $year$ model term.

# Model Fit Assessment

```{r echo=FALSE}
library("gridExtra")
anova(mod)
plot1 <- ggplot(mod, aes(sample = mod$residuals)) +
  stat_qq() + 
  stat_qq_line(linetype="dashed", color="blue") +
  labs(x = "Theoretical Quantiles",
       y = "Standardized Residuals",
       title = "Normal Q-Q Plot") +
  theme(text = element_text(size=t.size))
plot2 <- ggplot(df, aes(x = predicted.values, y = total, color=month)) + 
  geom_point() + geom_abline(intercept = 0, slope = 1, color="red") +
  colScale + labs(x = "Monthly Giving Predicted", 
                  y = "Monthly Giving Actual",
                  title = "Act. by Pred. Plot") +
  theme(text = element_text(size=t.size), 
        axis.text.x = element_text(angle=45, hjust=1)) +
  theme(legend.position="none")
plot3 <- ggplot(df, aes(x = predicted.values, y = residual.values, color = month)) +
  geom_point() + geom_hline(yintercept = 0, linetype="dashed", 
                            color = "blue") +
  colScale + labs(x = "Monthly Giving Predicted", y = "Residuals",
                  title = "Resid. by Pred. Plot") + 
  theme(text = element_text(size=t.size), 
        axis.text.x = element_text(angle=45, hjust=1)) +
  theme(legend.position="none")
plot4 <- ggplot(df, aes(x = as.numeric(row.names(df)), 
                         y = residual.values, 
                         color = month)) + 
  geom_point() + geom_hline(yintercept = 0, 
                            linetype="dashed", 
                            color = "blue") + 
  colScale +
  labs(x = "Row Number", 
       y = "Residuals",
       title = "Resid. by Row Plot") + 
  theme(text = element_text(size=t.size)) +
  theme(legend.position="none")
grid.arrange(plot2, plot3, plot4, plot1, ncol=2, nrow=2)
```

# Regressed Model Details

```{r echo=FALSE}
summary(mod)
```

This plot shows the regression coeffiecients capturing month-to-month variation...

```{r echo=FALSE}
mon.coef <- c(0, mod$coefficients[10:20])
names(mon.coef) <- month.name
#ggplot(data.frame(mon.coef), aes(seq_along(mon.coef),mon.coef))+geom_bar(stat = "identity")
ggplot(data = NULL, aes(x = factor(month.name, levels = month.name), 
                        y = mon.coef)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=round(mon.coef, digits = 2)),
            vjust=-0.25) +
  labs(x = "Month",
       y = "Month's Giving \"Adder\"",
       title = "Giving Variation By Month") +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

