---
title: "2018 Model Extension"
author: "Sean Davern"
date: "June 18, 2019"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
# Overview

This notebook extends the modeling done in 2013 to include data through much of 2018.

# Load Data
Loading cleaned and transformed giving data:
```{r}
df <- readRDS("../data/Cleaned and Transformed Giving Data.rds")
head(df)
```
# Columns to Accomodate Analysis
I'll add a few columns to enable asking analysis "what if's" and adjust for anomolies:
```{r}
library("dplyr", quietly=TRUE, warn.conflicts = FALSE)
df <- mutate(df, 
             excluded = FALSE, 
             removed.amount = 0, 
             original.total = total)
```

# Accommodating Unusual Gift
12/26/2016 had an unusual single gift skewing analyses.  Other large gifts have been recieved over this period but this gift was at least twice all others. I'm removing the amount of the unusual gift from the month's total.  Previously I excluded that month all-together, but I've commented that line out so no values are excluded.
```{r}
df$removed.amount[as.Date(df$week.ending) == "2016-12-25"] <- 20000
df$total <- df$original.total - df$removed.amount
#df$excluded[as.Date(df$week.ending) == "2016-12-25"] <- TRUE
```
Note: we're not excluding the 3 months consider "outliers" in the 2013 analysis.  Now with 9 years of history, those months don't appear as much to be outliers.

# Modeling
Regressing the R version of the 2013 model:
$$\textit{Monthly Giving} = a+b_{year}+c_{month}$$
```{r}
mod <- lm(
  formula = total ~
    year + month,
  data = df[df$excluded!=TRUE,]
)
```
...gives the following

# Model Predictions

```{r echo=FALSE}
library("magrittr")
suppressMessages(library("ggplot2", quietly = TRUE, warn.conflicts = FALSE))
library("RColorBrewer")
cbPalette <- brewer.pal(12,"Paired")  # Used http://colorbrewer2.org/
names(cbPalette) <- levels(month.name)
colScale <- scale_color_manual(name = "month", values = cbPalette)
t.size <- 9
df$week.ending <- as.Date(df$week.ending, "%Y-%m-%d", tz="UTC")
# The model's predicted values don't have values at the excluded months.
# Calculate those:
df <- mutate(df, predicted.values = predict(mod,df)) %>%
  mutate(residual.values = total - predicted.values )
ggplot(df, aes(x = as.Date(week.ending), y = total, color=month )) +
  geom_point() + geom_line(color='blue', 
                           data = df, 
                           aes(x=week.ending,y=predicted.values)) +
  colScale + labs(x = "Date", 
                  y = "Month's Total Giving ($)",
                  title = "Monthly Giving Model & Data") + 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") + 
  scale_y_continuous(breaks = seq(10000, 65000, by=5000),
                     labels = scales::comma) +
  theme(text = element_text(size=t.size),
        panel.grid.major = element_line(color = "azure2",
                                        linetype = "dashed",
                                        line), 
        axis.text.x = element_text(angle=45, hjust=1))
```
In this case it is far more clear that giving was increasing through 2016 which would be expected to improve the statistical significance of the $year$ model term over what was seen in the 2013 analysis.

\newpage
# Model Fit Assessment

```{r echo=FALSE}
library("gridExtra", quietly=TRUE, warn.conflicts = FALSE)
anova(mod)
plot1 <- ggplot(mod, aes(sample = mod$residuals)) +
  stat_qq() + 
  stat_qq_line(linetype="dashed", color="blue") +
  labs(x = "Theoretical Quantiles",
       y = "Standardized Residuals",
       title = "Normal Q-Q Plot") +
  theme(text = element_text(size=t.size))
plot2 <- ggplot(df, aes(x = predicted.values, y = total, color=month)) + 
  geom_point() + geom_abline(intercept = 0, slope = 1, color="red") +
  colScale + labs(x = "Monthly Giving Predicted", 
                  y = "Monthly Giving Actual",
                  title = "Act. by Pred. Plot") +
  theme(text = element_text(size=t.size), 
        axis.text.x = element_text(angle=45, hjust=1)) +
  theme(legend.position="none")
plot3 <- ggplot(df, aes(x = predicted.values, y = residual.values, color = month)) +
  geom_point() + geom_hline(yintercept = 0, linetype="dashed", 
                            color = "blue") +
  colScale + labs(x = "Monthly Giving Predicted", y = "Residuals",
                  title = "Resid. by Pred. Plot") + 
  theme(text = element_text(size=t.size), 
        axis.text.x = element_text(angle=45, hjust=1)) +
  theme(legend.position="none")
plot4 <- ggplot(df, aes(x = as.Date(week.ending), 
                         y = residual.values, 
                         color = month)) + 
  geom_point() + geom_hline(yintercept = 0, 
                            linetype="dashed", 
                            color = "blue") + 
  colScale +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") + 
  labs(x = "Date", 
       y = "Residuals",
       title = "Resid. by Date Plot") + 
  theme(text = element_text(size=t.size),
        legend.position="none", 
        axis.text.x = element_text(angle=45, hjust=1))
grid.arrange(plot2, plot3, plot4, plot1, ncol=2, nrow=2)
```

Both model terms are highly statistically significant.  The residuals seem nicely randomly scattered.  I don't see any particularly concerning patterns in the plots including model biases by month (dot color).  The quantile plot shows very normally distributed residuals.

\newpage
# Regressed Model Details

```{r echo=FALSE}
summary(mod)
```
# Model Generalization
We can generalize the model again, as described in the "Transfer of 2013 Analysis" Notebook, by shifting the coefficients to the average month and year like this:
$$
\textit{Monthly Giving} 
= \big[a+\overline{b_{year}}+\overline{c_{month}}\big]+
\big(b_{year}-\overline{b_{year}}\big)+
\big(c_{month}-\overline{c_{month}}\big)
$$
where $\overline{b_{year}}$ is the mean of $b_{year}$ coefficients, including $b_{2010}=0$, and $\overline{c_{month}}$ is the mean of the $c_{year}$ coefficients, including $c_{January}=0$. 

\newpage
# Monthly Variation
This plot shows the resulting $month$ regression coeffiecients shifted by $\overline{c_{month}}$ capturing month-to-month variation. ($(c_{month}-\overline{c_{month}})$)   The shape of the yearly repeated pattern in the prediction are the result of these terms.

```{r echo=FALSE}
mon.coef <- c(0, mod$coefficients[10:20])
mon.coef <- mon.coef - mean(mon.coef)
names(mon.coef) <- month.name
ggplot(data = NULL, aes(x = factor(month.name, levels = month.name), 
                        y = mon.coef)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=round(mon.coef, digits = 2)),
            vjust=-0.25) +
  labs(x = "Month",
       y = "Month's Giving \"Adder\"",
       title = "Giving Variation By Month") +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```
This plot is generally similar to Figure 13 in the 2013 Analysis with the exception of November. Since it is based on 9 years of Seed history, I'll judge that it is more representative of an average year.