---
title: "2019 Seed Financial Planning"
author: "Sean Davern"
date: "June 5, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
---

# Import of Data
The original data was provided by John Earling in a work book entitled 'Weekly PayPay & Tithes .xls' workbook.  The layout/format of that workbook was not conducive to easily loading into R [nor JMP originally] and so was transcribed into the workbook 'Giving Data.xlsx' and then read into R.

```{r extract, message=FALSE}
source("../code/0-Extract data from Excel.R")
df  # This is the data frame resulting from the import.
```
# Some Minor Data Validation
As a first validation I'll check that the weekly PayPal and offering amounts sum to the weekly totals $(paypal_i+{offering}_i\overset{?}=total_i)$, reporting only those that aren't equal:
```{r validate totals}
source("../code/1-Validate totals.R")
```
Ok, so December 2015 and 2016 seem to have totals greater than accounted for by the PayPal and offering amounts.  That's perhaps explainable by other end-of-year giving coming in another way.  However, the April 2017 discrepancy seems to be missing $500.  I'll need to look into that.

# Data Transformation

Aggregating the monthly totals and preparing to model month values...
```{r}
# Data transformation: Calculate monthly giving totals.
# Make Month a categorical variable with levels in the order that
# months occur in the year otherwise months are sorted alphabetically.
df$month <- factor(df$month, month.name)
# Aggregate the monthly Totals from giving.data in sums for each month.
MonthTotals <- 
  aggregate(df$total, by = list(df$month, df$year), FUN = sum)
# Exclude the months that don't have totals yet.
MonthTotals <- MonthTotals[complete.cases(MonthTotals), ]
# Extract only rows containing 'monthly.giving.families' data.
df <- df[!is.na(df$monthly.giving.families),]
# Now replace Totals (which were weekly totals) with calculated aggregates
df$total <- MonthTotals$x
# paypal & offering columns are now misleading (only week's value) so remove them.
df <- select(df, -paypal, -offering) 
```
Adding the number of giving Sundays in the month and the average giving each week per month...
```{r}
source("../code/NumOfGivenDayOfWeekInMonth.R")
# Calculate and add the columns SundaysInMonth with calculated values
# and MonthsGivingPerWeek
df <-  df %>%
  mutate(SundaysInMonth =
           NumOfGivenDayOfWeekInMonth(df$week.ending, "Sunday")) %>%
  mutate(MonthsGivingPerWeek = total / SundaysInMonth)
```
Enable modeling year as factor...
```{r}
# Make year a categorical variable so coefficients are easier to interpret.
df$year <- as.factor(df$year)
```
Save and print the resulting tibble:
```{r echo=TRUE}
# Write it as a csv:
write.csv(x = df,
          file = "../data/Cleaned and Transformed Giving Data.csv",
          row.names = FALSE)
# Save it as an R object
saveRDS(df, file = "../data/Cleaned and Transformed Giving Data.rds")
```





The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

