---
title: "2019 Seed Financial Planning"
author: "Sean Davern"
date: "June 5, 2019"
output: html_notebook
---

The original data was provided by John Earling in a work book entitled 'Weekly PayPay & Tithes .xls' workbook.  The layout/format of that workbook was not conducive to easily loading into R [nor JMP originally] and so was transcribed into the workbook 'Giving Data.xlsx' and then read into R.

```{r extract}
source("../code/0-Extract data from Excel.R")
df  # This is the data frame resulting from the import.
```

As a first validation I'll check that the weekly totals PayPal and offering amounts sum to the weekly total, reporting only those that don't:

```{r validate totals}
source("../code/1-Validate totals.R")
```

Ok, so December 2015 and 2016 seem to have totals greater than accounted for by the PayPal and offering amounts.  That's perhaps explainable by other giving coming in another way.  However, the April 2017 discrepancy seems to be missing $500.  I'll need to look into that.



The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

