---
title: "2019 Seed Financial Planning"
author: "Sean Davern"
date: "June 5, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
---

The original data was provided by John Earling in a work book entitled 'Weekly PayPay & Tithes .xls' workbook.  The layout/format of that workbook was not conducive to easily loading into R [nor JMP originally] and so was transcribed into the workbook 'Giving Data.xlsx' and then read into R.

```{r extract, message=FALSE}
source("../code/0-Extract data from Excel.R")
df  # This is the data frame resulting from the import.
```

As a first validation I'll check that the weekly PayPal and offering amounts sum to the weekly totals $(paypal_i+{offering}_i\overset{?}=total_i)$, reporting only those that aren't equal:
```{r validate totals}
source("../code/1-Validate totals.R")
```
Ok, so December 2015 and 2016 seem to have totals greater than accounted for by the PayPal and offering amounts.  That's perhaps explainable by other end-of-year giving coming in another way.  However, the April 2017 discrepancy seems to be missing $500.  I'll need to look into that.



The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

